---
title: Extending R's GTFS abilities with simple features
description: Two functions to easily transition from library(gtfsr) into library(sf) for tidyier transit analysis.
author: Nate Day
date: '2018-06-02'
slug: extending-gtfs-capabilities-with-parsing-into-simple-features
twitter_img: "images/gtfs_sf/gtfs_sf.jpeg"
categories:
  - Civic Data
tags:
  - sf
  - gtfsr
  - tidyverse
---

```{r setup, cache = F, include=F}
knitr::opts_chunk$set(error = TRUE, message = FALSE, warning = FALSE)
```

##### TLDR

[Two similar functions](https://gist.github.com/nathancday/d50afccb762174d402f54486d3d328b6) for building `sf` objects of the routes and stops within a `gtfs` objects.

## General Transit Feed Specification

Originally named Google Transit Feed Specification, GTFS was developed in [2005 by Chris Harrelson](https://en.wikipedia.org/wiki/General_Transit_Feed_Specification#History) (a Googler at the time), to make trip planning on civic transit systems easier.

GTFS format is a standardized set of comma separated `txt` files, with specific names and fields. To be flexible (i.e. cover a lot of public transit options), not all files/fields are required. Here is [a comprehensive list breakdown.](https://developers.google.com/transit/gtfs/reference/#file_requirements)

### library(gtfsr)

The [rOpenSci project's](https://ropensci.org/packages/) [`library(gtfsr)`](https://github.com/ropensci/gtfsr) is designed to make reading and visualizing GTFS data fast and easy. You can install the package from GitHub, via `devtools::install_github("ropensci/gtfsr")`.

#### `gtfsr::import_gtfs()`

This function does a great job handling the flexible GTFS structure. Here I'm reading in the local Charlottesville Area Transit's (CAT) GTFS feed.

The zipped data is available in [this Civic Innovation Day repo](https://github.com/Smart-Cville/CID-2018-Regional-Transit-Challenge).

```{r}
library(gtfsr)

cat <- gtfsr::import_gtfs("https://github.com/Smart-Cville/CID-2018-Regional-Transit-Challenge/blob/b6c714ec190f8843d6aa154fc74ed7be3bd5307f/data/2017_08_CharlottesvilleAreaTransit.zip?raw=true")
# lots of verbose warnings that I turned off for kniting porpoises
```

The function returns to us a special `S3 class` list object, `gtfs`, that has a data frame, for each of the available files.

```{r}
class(cat)
names(cat)
```

#### `gtfsr::map_gtfs()`

A high-level mapping function built with `library(leaflet)`.

In this specific case, attempting to plot all routes results in an error about missing the `agency_id` field. So my work around here is to explicitly specify all routes, which is easy via `%>%` dot.

```{r}
library(magrittr)

cat %>% map_gtfs() # errors out

# work around
cat %>% map_gtfs(., route_ids = .$routes_df$route_id,
                 route_colors = paste0("#", .$routes_df$route_color)) %>%
    # needed for web sharing only
    widgetframe::frameWidget() # isolates leaflet:css, to prevent messing with webpage:css
```

You can compare the `library(gtfsr)` results against the current CAT [app](http://catpublic.etaspot.net/) and [static map](http://www.charlottesville.org/home/showdocument?id=54576).

## Extending to `sf`

The `gtfsr` package does a great job reading and showing data, but if you want to join in additional data sources or customize plots beyond colors or routes, you are out of luck.

[Simple Features](https://en.wikipedia.org/wiki/Simple_Features) is making big waves in the [r-spatial](https://github.com/r-spatial) community since `library(sf)`(https://github.com/r-spatial/sf) introduced a new class of geometry enriched data frames.

#### Extracting routes into `routes_sf`

Because GTFS data is distributed across multiple files, we must pick apart the list of data frames to pull out the critical pieces.

In order to reconstruct a plot similar to `map_gtfs()`, we need:

1. key-value pairs from `cat$trips_df`
2. route meta information from `cat$routes_df`
3. shapes as Lon/lat coordinates from `cat$shapes_df`

* Caution: the following function is long *

```{r}
library(sf)
library(tidyverse)

gtfs_routes_sf <- function(gtfs) {
    
    ## gather key-values first ----
    
    # trips_df has route_id:shape_id
    shape_key <- gtfs$trips_df %>%
        select(route_id, shape_id) %>%
        unique()

    # routes_df has route_id:route_name
    route_key <- gtfs$routes_df %>%
        select(route_id, route_short_name) %>%
        mutate(route_short_name = paste("route", route_short_name)) %>%
        inner_join(shape_key)
    
    # check for colors :)
    if ( !is.null(gtfs$routes_df$route_color) ) { # extract if they exist
        route_key %<>% inner_join(select(gtfs$routes_df, route_color, route_id) )
    }
    else { # planB: build a pal from my favorite pallette 'd3'
        route_key %<>% mutate(route_color = rep(ggsci::pal_d3()(10),
                                                length.out = nrow(route_key)))
    }
    
    ## build the sf object ----
    
    # exctract lon/lat values as matrix to build linestrings for each "shape_id"
    sfc <- gtfs$shapes_df %>% # long data frame
        split(.$shape_id) %>% # list of shorted data framee, one per shape
        map(~ select(., shape_pt_lon, shape_pt_lat) %>% # order maters, lon-1st lat-2nd
            as.matrix %>% # coherce for st_linestrings happiness
            st_linestring) %>%
        st_sfc(crs = 4326) # bundle all shapes into a collection

    # add collection on and convert to sf
    unique(gtfs$shapes_df$shape_id) %>%
        sort() %>% # sort to match with names(sfc); split()'s factor-cohercion alpha sorts
        st_sf(shape_id = ., geometry = sfc) %>%
        inner_join(route_key)
    # plots fine ^^
    
    # st_sf(route_key, geometry = sfc)
    # doesn't plot ^^ and I can't explain it
}
```

OK that a lot of code, but does it actually work?

```{r}
cat_routes_sf <- gtfs_routes_sf(cat)

head(cat_routes_sf)
```

Yea it does.

Now all we need to do is viz it and since `library(leaflet)` already works out of the box with `sf` objects, it's cake.

```{r}
library(leaflet)

leaflet(cat_routes_sf) %>%
    addTiles() %>%
    addPolylines(color = ~paste0("#", route_color),
                 label = ~as.character(route_short_name)) %>%
    addLegend(colors = ~unique(paste0("#", route_color)),
              labels = ~unique(route_short_name)) %>%
    widgetframe::frameWidget()
```

#### Rinse and repeat for `stops_sf()`

The steps are slightly different but the pattern is the same (and it's still long).

```{r}
gtfs_stops_sf <- function(gtfs) {
    shape_key <- gtfs$trips_df %>%
        select(trip_id, route_id, shape_id) %>%
        unique()
    
    # stop_times_df also has stop sequence and arrive/depart time for specific stops
    stop_key <- gtfs$stop_times_df %>%
        select(trip_id, stop_id) %>%
        unique() %>%
        inner_join(shape_key) %>% # one stop is on multiple routes
        # need to pair down
        arrange(route_id) %>% # use route_id as tiebreaker (for now)
        group_by(stop_id) %>% # group_by() to stop_id 
        slice(1) # to slice() out the first row
    
    if ( !is.null(gtfs$routes_df$route_color) ) {
        stop_key %<>% inner_join(select(gtfs$routes_df, route_color, route_id)) }
    else {stop_key %<>% mutate(route_color = rep(ggsci::pal_d3()(10), length.out = nrow(route_key))) }
    
    stops_sfc <- gtfs$stops_df %>%
        split(.$stop_id) %>%
        map(~select(., stop_lon, stop_lat) %>%
                unlist() %>%
                st_point() ) %>% # point instead of linestring
        st_sfc()
    
    st_sf(stop_key, geometry = stops_sfc) %>%
        inner_join(gtfs$stops_df)
}
```

Now, make use and make `stops_sf` object.

```{r}
cat_stops_sf <- gtfs_stops_sf(cat)

head(cat_stops_sf)
```

Now we just have to combine our two layers, routes and stops.

To make a map that looks a lot like to the result from `map_gtfs()`, I'm gonna tweak a few of the defaults to `addCircleMarkers()`. 

```{r}
leaflet(cat_routes_sf) %>%
    addTiles() %>%
    addPolylines(color = ~paste0("#", route_color),
                 label = ~htmltools::htmlEscape(route_short_name)) %>%
    addCircleMarkers(data = cat_stops_sf, fillColor = ~paste0("#", route_color),
                     label = ~htmltools::htmlEscape(stop_name),
                     color = "black", radius = 5, weight = 3) %>%
    addLegend(colors = ~unique(paste0("#", route_color)),
              labels = ~unique(route_short_name)) %>%
    widgetframe::frameWidget()
```

👌

## Wrap up

Now we can build `sf` objects from `gtfs` objects at-will!

That gives us new plotting flexibility and makes incorporating new data more straight forward.

Looks like exploring transit patterns in R just got a little easier! Now if only all transit data was already cleaned into GTFS ...

```{r}
devtools::session_info()
```

