---
title: Extending R's GTFS abilities with simple features
description: Two functions to easily transition from library(gtfsr) into library(sf) for tidyier transit analysis.
author: Nate Day
date: '2018-05-31'
slug: extending-gtfs-capabilities-with-parsing-into-simple-features
twitter_img: "images/gtfs_sf/gtfs_sf.jpeg"
categories:
  - Civic Tech
tags:
  - sf
  - gtfsr
  - tidyverse
---

```{r setup, cache = F, include=F}
knitr::opts_chunk$set(error = TRUE, message = FALSE, warning = FALSE)
```

##### TLDR

A gist, [gtfs_sf.R](https://gist.github.com/nathancday/d50afccb762174d402f54486d3d328b6), with two functions for building `sf` objects from `gtfs` objects.

## General Transit Feed Specification

Originally named Google Transit Feed Specification, GTFS was developed in [2005 by Chris Harrelson](https://en.wikipedia.org/wiki/General_Transit_Feed_Specification#History) (a Googler at the time), to make trip planning on civic transit systems easier. By adopting a standardized system of files, local transit authorities could publish data that search tools could readily display for curious customers.

To be flexible, GTFS data only requires certain specific fields in specific files, but has optional fields to pretty much cover any kind of public transit you can think of. Here is a list of [which fields and files](https://developers.google.com/transit/gtfs/reference/#file_requirements) are must haves and which are just features.

### library(gtfsr)

The [rOpenSci project's](https://ropensci.org/packages/) [`library(gtfsr)`](https://github.com/ropensci/gtfsr) is designed for reading, validating and visualizing GTFS data at a high-level. You can install from GitHub, via `devtools::install_github("ropensci/gtfsr")`.

#### `gtfsr::import_gtfs()`

This function does a great job handling the flexible GTFS structure. Let's read in my local Charlottesville Area Transit's (CAT) GTFS feed as an example. I'm reading in from a url, but it also works with local paths, but you must specify `local = TRUE` as the second argument.

The zipped data is avaialbe in [this repo](https://github.com/Smart-Cville/CID-2018-Regional-Transit-Challenge).

```{r}
library(gtfsr)

cat <- gtfsr::import_gtfs("https://github.com/Smart-Cville/CID-2018-Regional-Transit-Challenge/blob/b6c714ec190f8843d6aa154fc74ed7be3bd5307f/data/2017_08_CharlottesvilleAreaTransit.zip?raw=true")
# lots of verbose warnings that I turned off for kniting porpoises
```

The function give us a special `S3 class` list object, called `gtfs`, that has a data frame, for each of the avaiable files.

```{r}
class(cat)
names(cat)
```

#### `gtfsr::map_gtfs()`

A high-level mapping function built with `library(leaflet)`. It gives us ability to filter data by routes/shapes and provide custom colors and labels. 

In this specific data case, attempting to plot all routes results in an error about missing the `agency_id` field. So my work around it specify all of the routes.

```{r}
library(magrittr)

cat %>% map_gtfs() # errors out

# work around
cat %>% map_gtfs(., route_ids = .$routes_df$route_id,
                 route_colors = paste0("#", .$routes_df$route_color)) %>%
    # needed for web sharing only
    widgetframe::frameWidget() # isolates leaflet:css, to prevent messing with webpage:css
```

You can compare the `library(gtfsr)` results against the current CAT [app](http://catpublic.etaspot.net/) and [static map](http://www.charlottesville.org/home/showdocument?id=54576).

## Extending to `sf`

The `gtfsr` package does a great job reading and showing data, but if you want to join in additional data sources or customize plots beyond colors or routes, you are out of luck.

[Simple Features](https://en.wikipedia.org/wiki/Simple_Features) have been making big waves in the [r-spatial](https://github.com/r-spatial) community via `library(sf)`(https://github.com/r-spatial/sf). These geometry enriched objects, behave a lot like data frames, which makes using R's tools like `dplyr` and `data.table` on them, straight forward.

I used `sf` objects in a earlier post to explore [local drug crime patterns.](/2017/12/30/drug-crime-analysis-with-library-sf/)

#### Converting routes into `routes_sf`

Because `cat` is a list with multiple data frames, we need to pick it apart into the needed pieces. To make a plot for routes, like `map_gtfs()`, we need:

1. key-value pairs from `cat$trips_df`
2. route meta information from `cat$routes_df`
3. shapes as lon/lat coordinates from `cat$shapes_df`

```{r}
library(sf)
library(tidyverse)

gtfs_routes_sf <- function(gtfs) {
    
    ## gather key-values first ----
    
    # trips_df has route_id:shape_id
    shape_key <- gtfs$trips_df %>%
        select(route_id, shape_id) %>%
        unique()

    # routes_df has route_id:route_name
    route_key <- gtfs$routes_df %>%
        select(route_id, route_short_name) %>%
        mutate(route_short_name = paste("route", route_short_name)) %>%
        inner_join(shape_key)
    
    # check for colors
    if ( !is.null(gtfs$routes_df$route_color) ) { # extract if possible
        route_key %<>% inner_join(select(gtfs$routes_df, route_color, route_id) )
    }
    else { # plan b is make some from my favorite pallette 'd3'
        route_key %<>% mutate(route_color = rep(ggsci::pal_d3()(10),
                                                length.out = nrow(route_key)))
    }
    
    ## build the sf object ----
    
    # exctract lon/lat values as matrix to build linestrings for each "shape_id"
    sfc <- gtfs$shapes_df %>% # long data frame
        split(.$shape_id) %>% # list of shorted data framee, one per shape
        map(~ select(., shape_pt_lon, shape_pt_lat) %>% # order maters, lon-1st lat-2nd
            as.matrix %>% # coherce for st_linestrings happiness
            st_linestring) %>%
        st_sfc(crs = 4326) # bundle all shapes into a collection

    # add collection on and convert to sf
    unique(gtfs$shapes_df$shape_id) %>%
        sort() %>% # sort to match with names(sfc); split()'s factor-cohercion alpha sorts
        st_sf(shape_id = ., geometry = sfc) %>%
        inner_join(route_key)
    # plots fine ^^
    
    # st_sf(route_key, geometry = sfc)
    # doesn't plot ^^ and I can't explain it
}
```

Ok that a lot of code, but does it actually work, I'll show you. 

```{r}
cat_routes_sf <- gtfs_routes_sf(cat)

head(cat_routes_sf)
```

And `library(leaflet)` works out of the box with `sf` objects, so that's awesome.

```{r}
library(leaflet)

leaflet(cat_routes_sf) %>%
    addTiles() %>%
    addPolylines(color = ~paste0("#", route_color),
                 label = ~htmltools::htmlEscape(route_short_name)) %>%
    addLegend(colors = ~unique(paste0("#", route_color)),
              labels = ~unique(route_short_name)) %>%
    widgetframe::frameWidget()
```

#### Rinse and repeat for `stops_sf()`

The steps are slighly different but the pattern is the same.

```{r}
gtfs_stops_sf <- function(gtfs) {
    shape_key <- gtfs$trips_df %>%
        select(trip_id, route_id, shape_id) %>%
        unique()
    
    # stop_times_df also has stop sequence and arrive/depart time for specific stops
    stop_key <- gtfs$stop_times_df %>%
        select(trip_id, stop_id) %>%
        unique() %>%
        inner_join(shape_key) %>% # one stop is on multiple routes
        # need to pair down
        arrange(route_id) %>% # use route_id as tiebreaker (for now)
        group_by(stop_id) %>% # group_by() to stop_id 
        slice(1) # to slice() out the first row
    
    if ( !is.null(gtfs$routes_df$route_color) ) {
        stop_key %<>% inner_join(select(gtfs$routes_df, route_color, route_id)) }
    else {stop_key %<>% mutate(route_color = rep(ggsci::pal_d3()(10), length.out = nrow(route_key))) }
    
    stops_sfc <- gtfs$stops_df %>%
        split(.$stop_id) %>%
        map(~select(., stop_lon, stop_lat) %>%
                unlist() %>%
                st_point() ) %>% # point instead of linestring
        st_sfc()
    
    st_sf(stop_key, geometry = stops_sfc) %>%
        inner_join(gtfs$stops_df)
}
```

Now the fun part, making our new shiny `stops_sf` object.

```{r}
cat_stops_sf <- gtfs_stops_sf(cat)

head(cat_stops_sf)
```

All we need to do now is add the new points on top of our routes map. To make a map that looks a lot like to the result from `map_gtfs()`, we'll tweak a few of the defaulys to `addCircleMarkers()`. 

```{r}
leaflet(cat_routes_sf) %>%
    addTiles() %>%
    addPolylines(color = ~paste0("#", route_color),
                 label = ~htmltools::htmlEscape(route_short_name)) %>%
    addCircleMarkers(data = cat_stops_sf, fillColor = ~paste0("#", route_color),
                     label = ~htmltools::htmlEscape(stop_name),
                     color = "black", radius = 5, weight = 3) %>%
    addLegend(colors = ~unique(paste0("#", route_color)),
              labels = ~unique(route_short_name)) %>%
    widgetframe::frameWidget()

```

## Wrap up

Now we can build `sf` objects from `gtfs` data at-will. This gives us new plotting flexibility with `leaflet` and makes incoroprating new data sources simple. Exploring transit system patterns just got a little easier, so off we go for more data!
