---
title: Maps with the new ggplot2 v3.0.0
author: Nate Day
date: '2018-08-04'
twitter_img: "images/post/2018-07-08-maps-with-the-new-ggplot2-v3-0-0_files/figure-html/unnamed-chunk-3-1.png"
slug: maps-with-the-new-ggplot2-v3-0-0
categories:
  - Civic Data
tags:
  - ggplot2
  - tidyverse
---

In honor of `ggplot2` [turning version 3 on CRAN](https://twitter.com/hadleywickham/status/1014511041609453568) I decided to make some maps of the 2010 census in Charlottesville, Virginia, to show off the new `geom_sf()` layer.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=F, message=F)
```
#### Packages

```{r r_libraries}
library(magrittr) # viva la %<>% 
library(tidyverse)
```

#### Theme prep

Universal settings for all of my ggplot's. These make typing easier and documents more consistant.

```{r theme_setup}
theme_set(cowplot::theme_map() +
              theme(panel.grid.major=element_line(colour="transparent")))
scale_fill_continuous <- function(...) ggplot2::scale_fill_continuous(..., type = "viridis")
```

#### Census Data

The tract level summary is available on the [city's ODP](http://opendata.charlottesville.org/datasets/us-census-tract-area-2010). But you could also use the [`tidycensus`]() package for another city's record.

```{r census_sensor}
tracts <- sf::read_sf("https://opendata.arcgis.com/datasets/63f965c73ddf46429befe1132f7f06e2_15.geojson")

tracts %<>% select(OBJECTID, area = AREA_, Population:Asian)
```

Let's look at that census data now and since we have `geom_sf()` thowing on aesthetics is easy. Here I'll use `tracts$Population` as `fill`.

```{r}
ggplot(tracts, aes(fill = Population)) +
    geom_sf()
```

Ok that's pretty freaking easy. No suprise that the city's largest population is around UVA's grounds and the Corner.

Lets' use our favorite facets with `geom_sf()` to explore the racial distribution of Whites, Blacks, American Indians, and Asians in the city.

```{r}
long_tracts <- tracts %>%
    gather("race", "pop", White:Asian)

ggplot(long_tracts, aes(fill = pop)) +
    geom_sf() +
    facet_wrap(~ race)
```

Damn, Charlottesville is really, really white.

To make a better viz about the non-white population patterns it would be nice to free the fill scales in each facet. And because this is `ggplot()` now, I can use on my favorite grid helper tool, `cowplot::plot_grid()`. Any alternatives, like [gridextra](), [egg]() or [patchwork](), are on the table too.

```{r}
long_tracts %>%
    split(.$race) %>%
    map(~ ggplot(., aes(fill = pop)) +
            geom_sf() +
            facet_wrap(~race) ) %>%
    cowplot::plot_grid(plotlist = .)
```

That's pretty fast and now we have a much better picture of each race's distribution in the city.

Being able to manipulate and make maps with the `tidyverse` is awesome. Working with `ggplot2` layers is straight forward and there already exist a ton of accessory packages, like `cowplot` that make formatting these ggobjects straight forward too!



