---
title: Make maps with Rstats
author: Nate
date: '2020-04-14'
slug: make-a-map-from-an-open-dataset
categories:
  - Open Data
  - Maps
tags: []
description: Three steps 
twitter_img: post/2019-01-21-is-the-weather-getting-wetter_files/its_gon_rain.jpg
---

## Making maps

Maps are cool and maps are useful, but I always here people say that making maps is hard. I think making maps is simpler today than ever before. So to show you what I mean, this post is how to map an open spatial dataset.

## Get it

Geojson formatted data is available in your local data portal, I promise.

I'm using Census tract data hosted on Charlottesville's OpenData portal.

```{r import}
library(sf)

tracts <- read_sf("https://opendata.arcgis.com/datasets/63f965c73ddf46429befe1132f7f06e2_15.geojson")
```

## Wrangle it

This step is the most variable depending on what you have to do. But this data set comes formatted pretty well, so I don't really have to do much.

I set up a numeric column with the fraction of residents who are black for my fill scale and a character column with text for a tooltip.

```{r}
library(tidyverse)

tracts <- tracts %>% 
  mutate(frac_black = round(Black / Population, 2),
         tooltip_text = glue::glue("{frac_black * 100} % of residents are black"))
```


## Map it

Leaflet has been around forever and is used everywhere on the web. There are lots of new mapping packages coming out for R, but I always go back to Leaflet becasue I think the easiest to use.

Leaflet requires scales be defined indendependly so I create a scale to fill with first. Then I pass in column names to polygon arguments, with the `~` syntax, which is very simlar to the `aes` syntax of `ggplot2`. 

```{r map}
library(leaflet)

black_pal <- colorNumeric("Greys", tracts$frac_black)

leaflet(tracts) %>% 
  addPolygons(fillColor = ~black_pal(frac_black),
              label = ~tooltip_text,
              color = "black", fillOpacity = 1) %>% 
  widgetframe::frameWidget(height = '400') # to fix blog CSS conflicts; you probably don't need this line
```

## Wrap up

Maps are super powerful for sharing info about communities. I hope after reading this you feel like you can make a map with Rstats.

